---
title: "Introduction to FlexAssays"
author: "Yuchen Li"
date: "Revised: 15 April, 2025"
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{FlexAssays-intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r style, echo=FALSE, results='asis'}
BiocStyle::markdown()
```

# Introduction

`FlexAssays` is a flexible and extensible container designed to store multiple 
matrix-like assays that may differ in their row and/or column dimensions. It is 
particularly well-suited for representing multimodal or multi-batch biological 
data, such as single-cell experiments with partially overlapping features or 
samples.

This vignette introduces the core design of the `FlexAssays` class, explains 
how to construct and manipulate `FlexAssays` objects, and demonstrates how 
flexibility rules control the mapping between assay dimensions and global 
identifiers.

# Anatomy of a `FlexAssays`

Before diving into construction and usage, it's helpful to understand the 
overall structure of a `FlexAssays` object. A `FlexAssays` object consists of 
the following components:

* **A list of matrix-like assays.** Each assay must have both row and column 
names. The actual storage formats are handled by the subclasses of 
`FlexAssays`, such as `SimpleFlexAssays`.

* `rowMap` and `colMap`: Logical mapping matrices that define how each assay 
aligns with the global row and column names. These two sparse matrices track 
which rows or columns of each assay are present in the shared global dimensional 
names.

* `assayClasses`: A character vector indicating the supported classes for 
assays. This is primarily used for validation. By default, it supports all 
classes (`NULL` or `character(0)`), but developers can restrict it to specific 
types e.g., `"matrix"` or `"dgCMatrix"`.

* `rowFlex` and `colFlex`: Flexibility attributes that determine whether new 
assays are allowed to modify the global row or column spaces. Valid rules 
include:
  * `"free"` (default): New names are permitted and will be appended to the 
global space.
  * `"fixed"`: All assays must share identical names.
  * `"bounded"`: New assays must not introduce new names.

```{r FlexAssays-image, echo=FALSE, fig.width=20, fig.height=10}
knitr::include_graphics("FlexAssays.svg", dpi = 300)
```

These components together provide a balance between structural consistency and 
flexibility across assays with potentially different dimension names. Detailed 
discussion of each component will follow in the subsequent sections.

# Construction of a `FlexAssays`

In this section, we demonstrate how to construct a `FlexAssays` object using 
different initialization approaches. We'll show how to initialize from 
matrix-like assays and how global row and column names are handled during 
construction.

Throughout this section, we assume both `rowFlex` and `colFlex` are set to 
`"free"`, which allows new assays to introduce previously unseen row or column 
names. Flexibility modes will be discussed in more detail in a later section.

```{r prepare, message=FALSE}
library(FlexAssays)

# Define a helper function to generate random sparse matrices
generate_sparse_matrix <- function(nrow, ncol, density = 0.75) {
  mat1 <- Matrix::rsparsematrix(nrow = nrow, ncol = ncol, density = density)
  rownames(mat1) <- sample(LETTERS, nrow(mat1))
  colnames(mat1) <- sample(letters, ncol(mat1))
  return(mat1)
}
```

## From multiple assays

We can initialize a `FlexAssays` from a list of matrix-like objects. Each matrix 
must have row and column names. The global row and column spaces will be 
determined based on the union of the row and column names across all assays.

```{r FlexAssays-from-list}
m1 <- generate_sparse_matrix(10, 6)
m2 <- generate_sparse_matrix(8, 13)

fa1 <- FlexAssays(list(m1, m2))
fa1
```

## From a single assay

Alternatively, we can create a `FlexAssays` by passing a single matrix-like 
object. In this case, `rownames` and `colnames` will be derived from the matrix 
itself.

```{r FlexAssays-from-matrix}
fa2 <- FlexAssays(m1)
fa2
```

## From an existing `FlexAssays`

If the input is already a `FlexAssays` object, the constructor will simply 
return the same object without modification.

```{r FlexAssays-from-FlexAssays}
fa3 <- FlexAssays(fa1)
fa3
```

## With initial `rownames` and `colnames`

In some scenarios, you may wish to explicitly define the global `rownames` and 
`colnames` at the time of construction, without relying on input assays.

```{r FlexAssays-from-dimnames}
fa4 <- FlexAssays(rownames = sample(LETTERS, 10), colnames = sample(letters, 8))
fa4
```

**Note** When specifying `rownames` and `colnames`, you must not provide any 
`assays` at the same time. This is because the global dimensions have already 
been explicitly defined, and any input matrix may introduce conflicts. 
Attempting to do so will raise an **error**:

```{r FlexAssays-from-dimnames-error, error = TRUE}
fa4 <- FlexAssays(
  m1,
  rownames = sample(LETTERS, 10), 
  colnames = sample(letters, 8)
)
```

# `rowMap` and `colMap`

In this section, we introduce two core components of a `FlexAssays` object: 
`rowMap` and `colMap`. These logical matrices record how the rows and columns 
of each assay align with the global row and column names, ensuring consistency 
across assays that may have different dimensions.

* `rowMap` is a logical `SVT_SparseMatrix` that defines how the rows of each 
assay align with the global row names. Each row of the `rowMap` corresponds to 
a global row name, and each column corresponds to an assay. The entries are 
`TRUE` if that particular row is present in the respective assay:

```{r rowMap}
rowMap(fa1)
```

`colMap` operates in the same manner, but for the global column space:

```{r colMap}
colMap(fa1)
```

We can use `rowMap()` and `colMap()` of a `FlexAssays` object to inspect how 
each assay's dimensions are mapped. However, these internal mappings are 
intended primarily for developers, and should not be modified directly by the 
end-users. Therefore, replacement methods like `rowMap<-` and `colMap<-` are 
not provided.

# Common operations on `FlexAssays`

## Global informations

We can inspect the overall structure of a `FlexAssays` object using several 
simple accessor functions. To check its dimensions, use `dim()`, which returns 
a two-element vector representing the number of rows and columns in the global 
space:

```{r dim}
dim(fa1)
```

To inspect the global row and column names, use `dimnames()`, `rownames()`, or 
`colnames()`:

```{r rownames}
rownames(fa1)
```

```{r colnames}
colnames(fa1)
```

To get basic information about the assays in a `FlexAssays` object, use 
`length()` and `names()` functions:

* `length()` returns the number of assays.
* `names()` returns a character vector of the assay names.

```{r length}
length(fa1)
```

```{r names}
names(fa1) <- c("aa", "bb")
names(fa1)
```

These methods allow you to quickly explore the global structure and metadata of 
a `FlexAssays` object. Exploring this information is often the first step 
before manipulating the internal assays.

## Fetching assays

We can retrieve the list of assays from a `FlexAssays` object using 
`assays()`. This function should always return a `SimpleList` containing all 
assays. By default, the returned assays will include dimension names mapped 
from the global row and column spaces.

```{r assays}
assays(fa1)
```

To access an individual assay, index into the list:

```{r assays-subset1}
assays(fa1)[[1]]
```

```{r assays-subset2}
assays(fa1)[[2]]
```

**Note**: The dimensional orders of the fetched `assays(fa1)[[2]]` may 
differ from the original matrix `m2`, but all corresponding entries remain 
correct. Internally, all assays are reordered to match the global dimension 
names defined in `rowMap` and `colMap`.

```{r original-matrix}
m2
```

For memory efficiency, the assays stored internally do not retain their own row 
and column names. Since the dimension names are already managed by `rowMap` and 
`colMap`, they are restored on demand. We can suppress this behavior by setting 
`withDimnames = FALSE`:

```{r assays-without-dimnames}
assays(fa1, withDimnames = FALSE)[[2]]
```

To fetch a single assay directly, we may also use the wrapper function 
`assay()`, which applies only the dimension name restoration:

```{r assay}
assay(fa1, 2)
```

In addition, `FlexAssays` supports list-like extraction via `[[` and `$`, just 
like `SimpleList` or similar list-based containers:

```{r list-subset}
fa1[[1]]
```

```{r list-subset-name}
fa1[['aa']]
```

The `$` operator can be used to extract assays by name, but it only works when 
the `FlexAssays` object has non-empty assay names:

```{r list-subset-dollar}
fa1$bb
```

## Modify assays

Modifying the assays include operations like adding new assays, replacing 
existing ones, and removing assays from a `FlexAssays`. These operations are 
subject to the flexibility rules defined by `rowFlex` and `colFlex`. In this 
section, we continue to assume both are set to `"free"`.

We can add a new assay to an existing `FlexAssays` by assigning it to the 
object using `[[<-`. This operation ensures that the assay's row and column 
names are aligned to the global dimension space.

```{r set-new-assay}
# Add a new assay to the FlexAssays object
m3 <- generate_sparse_matrix(10, 8)
fa1[["m3"]] <- m3
fa1
```

Alternatively, we can use the `$<-` operator to add a new assay by name:

```{r set-new-assay-dollar}
m4 <- generate_sparse_matrix(4, 8)
fa1$m4 <- m4
fa1
```

To replace an existing assay, assign a new assay to the corresponding name or 
index with `[[<-` or `$<-`:

```{r set-new-assay-list}
m3 <- generate_sparse_matrix(6, 5)
m4 <- generate_sparse_matrix(4, 8)
fa1[[3]] <- m3
fa1$m4 <- m4
fa1
```

To remove an existing assay, assign `NULL` to its name or index:

```{r remove-assay}
fa1$m4 <- NULL
fa1
```

We can also modify the content of an existing assay directly using standard 
indexing:

```{r modify-assay}
fa1$m3[1, ] <- 1
fa1$m3
```

## Renaming rows and columns

We can rename the global row and column names of a `FlexAssays` object using 
the standard `rownames()<-` and `colnames()<-` functions.

```{r modify-dimnames}
rownames(fa1) <- paste0("AAA_", rownames(fa1))
colnames(fa1) <- paste0("BBB_", colnames(fa1))
fa1
```

This operation updates the row and column names of all assays and automatically 
reflects the changes across all assays, ensuring consistent alignment 
throughout the `FlexAssays` object.

```{r update-internal-assay}
fa1$m3
```

## Subsetting `FlexAssays`

We can subset a `FlexAssays` object using the standard matrix-like indexing 
syntax, e.g., `[i, j]`. When subsetting, the global row and column names are 
respected, and the assays are subsetted accordingly:

```{r subset-index}
fa1[1:4, 1:5]
```

Both logical and character indexing are supported:

```{r subset-dimnames}
fa1[rownames(fa1)[1:4], colnames(fa1)[1:5]]
```

```{r subset-logical}
fa1[rownames(fa1) %in% paste0("AAA_", LETTERS[1:10]), ]
```

By default, assays that become empty after subsetting are automatically 
dropped. For example, if we subset out all rows used by `m3`, the resulting 
`FlexAssays` object will exclude `m3`:

```{r subset-drop}
# This will make 'm3' empty
fa2 <- fa1[!rownames(fa1) %in% rownames(fa1$m3), ]
fa2
```

To retain empty assays after subsetting, set `drop = FALSE`:

```{r subset-nodrop}
fa2 <- fa1[!rownames(fa1) %in% rownames(fa1$m3), , drop = FALSE]
fa2
```

## Clean unused dimension names

When an assay is removed, some global row or column names may no longer be used 
by any remaining assay. These unused dimension names are retained by default, 
but can be cleaned up using `cleanDimMaps()`.

For example, after removing assay `m3`, some global row names may become unused:

```{r rowmap-after-removing-assay}
fa1$m3 <- NULL
rowMap(fa1)
```

Calling `cleanDimMaps(fa1)` will remove any global rows or columns that are no 
longer mapped to any assay:

```{r clean-rowmap}
fa1 <- cleanDimMaps(fa1)
rowMap(fa1)
```

# Control of flexibility

## `rowFlex` and `colFlex`

In this section, we explain the roles of the `rowFlex` and `colFlex` 
attributes, which control how assays interact with the global row and column 
dimensions in a `FlexAssays` object.

Flexibility rules must be specified at the time of object initialization. We 
can access these rules using the `rowFlex()` and `colFlex()` getters:

```{r rowflex}
fa1 <- FlexAssays(list(m1, m2))
rowFlex(fa1)
```

By default, the flexibility mode is `"free"`, meaning the global row or column 
names can be extended automatically. For example, we first check the original 
row names of `fa1`:

```{r check-rownames-for-rowflex}
rownames(fa1)
```

Now we add a new assay `m3` that contains new row names not present in `fa1`. 
In `"free"` mode, these names will be added to the global row space:

```{r extend-rows-for-free}
m3 <- generate_sparse_matrix(6, 12)
rownames(m3)[1:3] <- setdiff(LETTERS, c(rownames(fa1), rownames(m3)))[1:3]
fa1$m3 <- m3
rownames(fa1)
```

In the next example, we will set `rowFlex` to `"bounded"` when `fa1` is 
created. Since `m3` contains row names not included in `fa1`, an error will be 
raised when `m3` is added.

```{r extend-rows-for-bounded-will-fail, error = TRUE}
fa1 <- FlexAssays(list(m1, m2), rowFlex = "bounded")
fa1$m3 <- m3
```

This constraint also applies to empty `FlexAssays` objects initialized with 
explicit global row names:

```{r predefined-rowflex, error = TRUE}
# Create an empty FlexAssays with pre-defined row names
fa2 <- FlexAssays(rownames = LETTERS[1:6], rowFlex = "bounded")

rownames(m3) <- LETTERS[4:9]
fa2$m3 <- m3
```

Therefore, we must make sure that `m3` contains valid row names before it is 
added. However, since the `colFlex` is not explicitly constrained, `fa2` can 
still automatically extend its column names.

```{r}
rownames(m3) <- rownames(fa2)[1:nrow(m3)]
fa2$m3 <- m3
fa2
```

`colFlex` works in the same way as `rowFlex`. For example, if `colFlex` is set 
to `"bounded"`, adding an assay with column names outside the predefined set 
will also result in an error:

```{r, error = TRUE}
# Create an empty FlexAssays with pre-defined row/column names
fa2 <- FlexAssays(
  rownames = LETTERS[1:6], rowFlex = "bounded",
  colnames = letters[1:15], colFlex = "bounded"
)
rownames(m3) <- rownames(fa2)[1:nrow(m3)]
colnames(m3) <- letters[10:21]
fa2$m3 <- m3
```

The new assay can only be added if its column names are within the allowed set:

```{r}
colnames(m3) <- letters[1:ncol(m3)]
fa2$m3 <- m3
fa2
```

For `"bounded"` flexibility, the valid dimension names are determined by the 
union of the initial assays, unless explicitly defined using `rownames = ...` 
or `colnames = ...` at construction. This mode is useful for scenarios where 
each assay may represent a subset of the full dataset, such as raw counts for 
all genes and normalized values for only highly variable genes in a typical 
single-cell analysis workflow.

If you want all assays to use exactly the same set of row or column names, use 
the `"fixed"` mode. For example, when storing RNA and ATAC matrices from the 
same cells, the column names should be identical.

When `colFlex = "fixed"` is used, the constructor will validate whether all 
input assays share the same set of column names:

```{r, error = TRUE}
m1 <- generate_sparse_matrix(6, 7)
m2 <- generate_sparse_matrix(4, 7)
colnames(m2) <- setdiff(letters, colnames(m1))[1:ncol(m2)]

# Error
FlexAssays(list(m1, m2), colFlex = "fixed")
```

It succeeds when all column names are identical:

```{r}
colnames(m2) <- colnames(m1)
fa1 <- FlexAssays(list(m1, m2), colFlex = "fixed")
fa1
```

In `"fixed"` mode, you also cannot add assays that differ in dimension names:

```{r, error = TRUE}
m3 <- generate_sparse_matrix(5, 7)
colnames(m3)[1:3] <- setdiff(letters, colnames(fa1))[1:3]
fa1$m3 <- m3
```

New assays must exactly match the global column names to be accepted:

```{r}
colnames(m3) <- sample(colnames(fa1))
fa1$m3 <- m3
fa1
```

## `assayClasses`

In addition to controlling global dimension names, `FlexAssays` can also 
enforce restrictions on the class of internal assays using the `assayClasses` 
argument. This is particularly useful when storing matrices that must support 
specific algorithms or backends, such as `DelayedArray`, `CsparseMatrix`, or 
other sparse matrix representations.

The `assayClasses` constraint is set at initialization and cannot be modified 
afterward. For example, here we construct a `FlexAssays` object that only 
accepts assays of class `"CsparseMatrix"` or `"RsparseMatrix"`:

```{r, error = TRUE}
m1 <- generate_sparse_matrix(4, 6)
m2 <- generate_sparse_matrix(5, 3)
m3 <- generate_sparse_matrix(8, 5)

m2 <- as(m2, "RsparseMatrix")
m3 <- as.matrix(m3)  ## Invalid assayClasses

assayClasses <- c("CsparseMatrix", "RsparseMatrix")
fa1 <- FlexAssays(list(m1, m2, m3), assayClasses = assayClasses)
```

This restriction also applies when adding new assays to an existing object:

```{r, error = TRUE}
fa1 <- FlexAssays(list(m1, m2), assayClasses = assayClasses)
fa1$m3 <- m3
```

# Session information

```{r}
sessionInfo()
```
